
---
title: HMEC ECM Protein Analysis 
author: Mark Dane

output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    source_code: embed
---


```{r "setup", include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
knitr::opts_knit$set(root.dir = "/data/share/dane/HMEC_ECMs")

```


```{r LoadAndCleanData, cache=TRUE}
#Author: Mark Dane, copyright 2017
library(ggplot2)
library(scales)
library(data.table)
library(MEMA)
library(grid)
library(knitr)
library(gplots)
library(RColorBrewer)
library(DT)
library(d3heatmap)
library(plotly)
library(corrplot)
library(Rtsne)
library(synapseClient)
library(NMF)

#Setup colors for Barcode and text in all heatmaps
selDark2 <- colorRampPalette(brewer.pal(8,"Dark2"))
plateCol <- selDark2(8)
hmcols<-colorRampPalette(c("blue","white","red"))(16)

#Load the latest level 3 SS4 HMEC datasets from Synapse
synapseLogin()
HMEC122L_l3 <- "syn7121240"
HMEC240L_l3 <- "syn7121244"
#Get and clean the datasets
l3L <- lapply(c(HMEC240L_l3,HMEC122L_l3), function(x){
  synGet(x) %>%
    getFileLocation() %>%
    fread(.,showProgress=FALSE) %>%
    filter(grepl("FBS",Ligand)) #Filter to only the FBS wells
})
#Combine into one data.table
l3 <- rbindlist(l3L)
#Remove FBS plate suffixes
l3$Ligand <- gsub("FBS_P.*","FBS",l3$Ligand)


```


###Experiment Overview


```{r Content Layout, fig.height=5, fig.width=5.5}

#Select the A row wells and delete the blanks
setkey(l3,Well)
dt <- unique(l3[grep("A",unique(l3$Well),value=TRUE),list(ArrayRow,ArrayColumn,ECMp)])
setkey(dt,ECMp)
dt <- dt[!grepl("blank|PBS",dt$ECMp),]

p <- ggplot(data=dt[!grepl("COL1",dt$ECMp),],aes(x = ArrayColumn, y = ArrayRow, fill=ECMp))+
  geom_point(shape=21, size = 2.7)+
  theme(legend.text = element_text(size = rel(.5)),legend.title=element_text(size = rel(.5)),plot.title=element_text(size = rel(.8)))+
  scale_y_reverse()+
  xlab("")+ylab("")+
  ggtitle("MEMA ECM Protein Layout")+
  theme(legend.text=element_text(size = rel(.5)),legend.title=element_text(size = rel(.8)))
p <- p+geom_point(data = dt[grepl("COL1",dt$ECMp),], shape=24, fill="black")
print(p)

#Add SCC pseudoimages


#Well level COL1 Spot Cell Count Boxplots
dt <- l3[grepl("COL1",ECMp),]
p <- ggplot(dt, aes(x=factor(Barcode), y=Spot_PA_SpotCellCount, colour=CellLine))+
  geom_boxplot(outlier.size = .5)+
  xlab("Well Plate Barcode")+ylab("Spot Cell Count")+
  ggtitle("COL1 Spot Cell Count by Plate")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0, hjust=0.5, size=rel(1)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
print(p)

#Well level Proliferation Boxplots
p <- ggplot(dt, aes(x=factor(Barcode), y=Nuclei_PA_Gated_EdUPositiveProportion, colour=CellLine))+
  geom_boxplot(outlier.size = .5)+
  xlab("Well Plate Barcode")+ylab("Proliferation Proportion")+
  ggtitle("COL1 Proliferation Proportion by Plate")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0, hjust=0.5, size=rel(1)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
print(p)

```

***
This project is a data mining exploration of the signals modulated by the ECM (Extracellular Matrix) proteins in the two MEMA (Microenvironment Microarray) HMEC cell line datasets from Oregon Health and Science University's MEP-LINCS project https://www.synapse.org/#!Synapse:syn2862345/wiki/394513.

There are eight replicate MEMAs in each cell line dataset, each in a different plate. 

All eight plates of a cell line were processed as a batch.

This analysis is based on immunofluorescent cell-level data that has been summarized to the spot and ECM protein levels. 

#####MEMA Layout
All MEMAs in the experiment are in separate wells and have the same design of 48 ECM proteins spotted in 35 rows and 20 columns. The proteins are randomly assigned to spots in the top 30 rows. Rows 31-35 are replicates of rows 1-5. There is a higher number of COL1 spots (shown as black triangles) throughout the array that are control spots. The upper left and bottom right corners of each MEMA are image fiducials in the 488nm channel and there are four blank spots for checking orientation in all channels.

#####Boxplots
The boxplots show the spot cell count and proliferation values for the 99 COL1 controls spots in each well. These wells are replicates within each cell line.

###Boxplots of Raw and Normalized cell count and proliferation

```{r boxplotsbyECMp, fig.height=5}

dt <- l3[!grepl("632|633|635|636|638|655",Barcode),]
normToCOL1Median <- function(dt){
  m <- numericMedian(unlist(dt[dt$ECMp=="COL1",1]))
  if(m==0)stop(paste("Can't normalize when median equals 0",unique(dt$Barcode)))
  r <- dt[,1]-m
}
#Normalize some signals
dt <- dt[,Spot_PA_SpotCellCountLog2NormCOL1Median := normToCOL1Median(.SD), .SDcols=c("Spot_PA_SpotCellCountLog2","ECMp","Barcode"),by=c("Barcode")]
dt <- dt[,Nuclei_PA_Gated_EdUPositiveProportionLogitNormCOL1Median := normToCOL1Median(.SD), .SDcols=c("Nuclei_PA_Gated_EdUPositiveProportionLogit","ECMp","Barcode"),by=c("Barcode")]
p <- ggplot(dt, aes(x = ECMp, y = 2^Spot_PA_SpotCellCountLog2, fill=CellLine))+
  geom_boxplot()+
  ggtitle(paste("Raw Spot Cell Count by ECM Protein and Cell Line"))+
  xlab("ECM Protein")+ylab("Cells per Spot")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.8)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
suppressWarnings(print(p))


p <- ggplot(dt, aes(x = ECMp, y = 2^Spot_PA_SpotCellCountLog2NormCOL1Median, fill=CellLine))+
  geom_boxplot()+
  #coord_cartesian(ylim=c(0,5))+
  ggtitle(paste("Normalized Spot Cell Count by ECM Protein and Cell Line"))+
  xlab("ECM Protein")+ylab("Normalized Cell Count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.8)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
#p <- p + geom_hline(yintercept = 1, colour="blue", alpha=.5)

suppressWarnings(print(p))

p <- ggplot(dt, aes(x = ECMp, y = btLogit(Nuclei_PA_Gated_EdUPositiveProportionLogit), fill=CellLine))+
  geom_boxplot()+
  ggtitle(paste("Raw Proliferation by ECM Protein and Cell Line"))+
  xlab("ECM Protein")+ylab("Proliferation Proportion")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.8)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
suppressWarnings(print(p))

p <- ggplot(dt, aes(x = ECMp, y = btLogit(Nuclei_PA_Gated_EdUPositiveProportionLogitNormCOL1Median), fill=CellLine))+
  geom_boxplot()+
  #coord_cartesian(ylim=c(0,5))+
  ggtitle(paste("Normalized Proliferation by ECM Protein and Cell Line"))+
  xlab("ECM Protein")+ylab("Normalized Proliferation")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=rel(.8)), axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1, size=rel(1)), plot.title = element_text(size = rel(1)),legend.text=element_text(size = rel(.8)),legend.title=element_text(size = rel(.8)))
#p <- p + geom_hline(yintercept = 1, colour="blue", alpha=.5)

suppressWarnings(print(p))


```

***
These boxplots group the values across the eight plates of a cell line then stratify them by ECM protein. 

#####Normalization Method
Each value is normalized to the COL1 spots in the same MEMA by log or logit transforming, subtracting the median COL1 value, then backtransforming the results.

###Heatmaps of Raw and Normalized cell count and proliferation

```{r heatmapsOfReplicates, fig.height=5}

#dt <- dt[!grepl("NID1|ELN",ECMp),]
dtc <- dcast(dt,ECMp~Barcode+CellLine,fun.aggregate = median, value.var = "Spot_PA_SpotCellCountLog2", fill=NA)
dtcRowNames <- dtc$ECMp
dtc <- dtc[,ECMp:=NULL]
dtcMatrix <- as.matrix(dtc, dimnames=list(dtcRowNames,colnames(dtc)))
#Compress the tails to show more variation within the distributions
dtcMatrix[dtcMatrix < quantile(dtcMatrix,probs = .05)] <- quantile(dtcMatrix,probs = .05)
dtcMatrix[dtcMatrix > quantile(dtcMatrix,probs = .95)] <- quantile(dtcMatrix,probs = .95)
rownames(dtcMatrix)<-dtcRowNames
aheatmap(dtcMatrix, col=hmcols, cexRow=.6, cexCol=.5,
          main=paste("Raw Spot Cell Count Replicates"),
        annCol = list(CellLine=as.factor(gsub(".*_","",colnames(dtcMatrix)))))

dtc <- dcast(dt,ECMp~Barcode+CellLine,fun.aggregate = median, value.var = "Spot_PA_SpotCellCountLog2NormCOL1Median", fill=NA)
dtcRowNames <- dtc$ECMp
dtc <- dtc[,ECMp:=NULL]
dtcMatrix <- as.matrix(dtc, dimnames=list(dtcRowNames,colnames(dtc)))
#Compress the tails to show more variation within the distributions
dtcMatrix[dtcMatrix < quantile(dtcMatrix,probs = .05)] <- quantile(dtcMatrix,probs = .05)
dtcMatrix[dtcMatrix > quantile(dtcMatrix,probs = .95)] <- quantile(dtcMatrix,probs = .95)
rownames(dtcMatrix)<-dtcRowNames
aheatmap(dtcMatrix, col=hmcols, cexRow=.6, cexCol=.5,
          main=paste("Normalized Spot Cell CountReplicates"),
        annCol = list(CellLine=as.factor(gsub(".*_","",colnames(dtcMatrix)))))


dtc <- dcast(dt,ECMp~Barcode+CellLine,fun.aggregate = median, value.var = "Nuclei_PA_Gated_EdUPositiveProportion", fill=NA)
dtcRowNames <- dtc$ECMp
dtc <- dtc[,ECMp:=NULL]
dtcMatrix <- as.matrix(dtc, dimnames=list(dtcRowNames,colnames(dtc)))
#Compress the tails to show more variation within the distributions
dtcMatrix[dtcMatrix < quantile(dtcMatrix,probs = .05)] <- quantile(dtcMatrix,probs = .05)
dtcMatrix[dtcMatrix > quantile(dtcMatrix,probs = .95)] <- quantile(dtcMatrix,probs = .95)
rownames(dtcMatrix)<-dtcRowNames
aheatmap(dtcMatrix, col=hmcols, cexRow=.6, cexCol=.5,
          main=paste("Raw Proliferation Replicates"),
        annCol = list(CellLine=as.factor(gsub(".*_","",colnames(dtcMatrix)))))


dtc <- dcast(dt,ECMp~Barcode+CellLine,fun.aggregate = median, value.var = "Nuclei_PA_Gated_EdUPositiveProportionLogitNormCOL1Median", fill=NA)
dtcRowNames <- dtc$ECMp
dtc <- dtc[,ECMp:=NULL]
dtcMatrix <- as.matrix(dtc)
#Compress the tails to show more variation within the distributions
dtcMatrix[dtcMatrix < quantile(dtcMatrix,probs = .05)] <- quantile(dtcMatrix,probs = .05)
dtcMatrix[dtcMatrix > quantile(dtcMatrix,probs = .95)] <- quantile(dtcMatrix,probs = .95)
rownames(dtcMatrix)<-dtcRowNames
colnames(dtcMatrix) <- gsub("_.*","",colnames(dtc))
aheatmap(dtcMatrix, col=hmcols, cexRow=.6, cexCol=.5,
          main=paste("Normalized Proliferation Replicates"),
        annCol = list(CellLine=as.factor(gsub(".*_","",colnames(dtc)))))


```

***
These heatmaps are of the raw and normalized cell count and proliferation signals. The outlier values are squished towards the centers to show greater variation in the heatmaps.


###PCA Plots of many signals

